---
title: A wedding social network
author: ~
date: '2019-05-18'
slug: a-wedding-social-network
categories: []
tags: []
image:
  caption: ''
  focal_point: ''
---
<!-- intro -->
Wedding season will soon be in full swing. During cocktail hour or over the salad course at dinner, a common question that guests ask one another is how they know the happy couple. The answers that come up  are localized to *particular individuals*, which means it's difficult to get a big picture understanding of how *entire groups of people* converge to celebrate the start of a marriage.

Thinking about this shortly before our own wedding last April, my then-fiancee (now wife :smiley:) and I conducted a social network analysis of the guests who would celebrate with us on our special day. So we gathered a list of our 134 guests and placed their names down column A and across row 1 of an Excel spreadsheet, creating a 134 x 134 adjacency matrix. Each entry in the matrix represents if two guests had previously met in any capacity (0 = no, 1 = yes). It took us a couple hours spread over a few days to fill in the matrix, but we ended up with the following. Guests' names are row and column labels. White squares mean the two guests hadn't met and black  squares mean they had met. Every square along the diagonal is black since guests have of course met themselves.

<!-- data prep -->

```{r include = FALSE}
# packages needed
if (!require("tidyverse")) {install.packages("tidyverse")}; require("tidyverse")
if (!require("corrplot")) {install.packages("corrplot")}; require("corrplot")
if (!require("qgraph")) {install.packages("qgraph")}; require("qgraph")

# load data
read_csv(file = "subjective.network.csv") -> admat_lower # adjacency matrix, lower half only
read_csv(file = "groups.csv") -> groups                  # organize guests into groups

# labels for groups and shapes
groups %>%
  mutate(group_label = case_when(group == 1 ~ "Kate's family", 
                                 group == 2 ~ "Kate's co-workers", 
                                 group == 3 ~ "Kate's college friends", 
                                 group == 4 ~ "Kate's HS friends", 
                                 group == 5 ~ "New Orleans friends", 
                                 group == 6 ~ "Jack's HS friends", 
                                 group == 7 ~ "Jack's college friends", 
                                 group == 8 ~ "Jack's grad school friends", 
                                 group == 9 ~ "Jack's family", 
                                 group == 10 ~ "Newlyweds"), 
         shape = ifelse(test = name == "Jack Cao" | 
                          name == "Kate Leist", 
                        yes = "heart", 
                        no = "circle")) -> groups
# convert admat_lower to matrix
admat_lower %>%
  select(2:ncol(admat_lower)) %>%
  as.matrix() -> admat_lower

# create full adjacency matrix
  # 134 x 134 identity matrix
admat_full <- diag(nrow(admat_lower))

  # add column and row labels, first names only
word(string = colnames(admat_lower), start = 1, sep = " ") -> colnames(admat_full) -> rownames(admat_full)

  # make lower half admat_full the same as lower half of admat_lower
admat_full[lower.tri(admat_full)] <- admat_lower[lower.tri(admat_lower)]

  # reflect lower half to upper half
admat_full + t(admat_full) -> admat_full

  # make diag equal to 1
diag(admat_full) <- 1

  # check if symmetric
isSymmetric(admat_full)
```

<!-- adjacency matrix -->

```{r echo = FALSE, fig.cap = "Adjaceny matrix, organized by each guest's loading onto the first principal component"}

corrplot::corrplot(corr = admat_full, 
                   method = "square", 
                   type = "full",
                   diag = TRUE,
                   order = "FPC",
                   tl.cex = 0.25,
                   tl.col = "black",
                   cl.pos = "n",
                   col = colorRampPalette(c("green","white","black"))(200))
  
```

<!-- adjacency matrix discussion -->

A few things pop out about the adjacency matrix.
  
* First, it's symmetrical. So instead of filling in all $134^{2}$ entries, we just filled in a little under half, $\frac{134^{2} - 134}{2}$.  
* Already there's strucutre. In the upper left corner, there's the biggest cluster of people who've all met before, which is Kate's extended family. And closer to the middle is a smaller but still sizeable cluster of people who've also met before, which are people in my lab.  
* The matrix is relatively sparse. Most squares are white, so lots of guests hadn't met other guests before. This makes sense since we invited people from different parts of our lives: family from China and Chicago, classmates from New Jersey, and colleages in Boston.

<!-- social network -->

```{r echo = FALSE}

qgraph(input = admat_full,                               ## Adjacency matrix
       layout = "spring",
       groups = as.factor(groups$group_label),    
       palette = "ggplot2",
       labels = FALSE,
       vsize = 1.5,                                      ## Node size
       shape = as.character(groups$shape),               ## Node shape
       border.color = "black",                           ## Border color of nodes
       unCol = "lightgrey",                              ## Edge color
       legend = TRUE, 
       legend.cex = 0.5)

```








